---
title: "A Statistical Test of the Symmetry of the Stock Returns Distributions"
author: "Dr. Hossain Pezeshki"
date: '2015-07-08'
output: html_document
css: style.css
bibliography: bibliography.bib
---
****

# Prompt and Summary #
A recent online [article](http://www.bloombergview.com/articles/2015-06-25/fear-factor-drives-the-stock-market-s-highs-and-lows) [@skewbbg] reports the existence of negative skewness in the
distribution of (some) stock returns, and we decided to test this assertion.
That is, we decided to test stock returns data for symmetry.
Commonly, the statistical [skewness](#skewreview) is used for this purpose. However, in order for
skewness to be meaningful at all, the distribution must have finite third order moments.
Furthermore the test of significance of sample skewness requires the existence of bounded moments
of sixth order. Thus, the fact that the observed return distributions are often
heavy-tailed [page 87 @ruppertfinance] makes skewness inappropriate
for the task at hand, and a different test is needed. We [propose such a test](#newtest)
after [reviewing the skewness metric](#skewreview).

Although our primary focus is finance, our discussion is broadly applicable to large sample statistics.

***

## Skewness and its requirement of bounded moments  <a name="skewreview"></a> ##
Let $x = [x_1, x_2,\ldots,
x_N]$ be a sample from a population governed by the pdf $f_X()$. Then, 
if $\int_{\mathbb{R}} \left|r \right|^3 f_X(r)\,dr < \infty$, define
the population skewness $\gamma_3$ by
$$
\gamma_3 \stackrel{\Delta}{=}
E \left(  
\dfrac{X - \mu_1}{\sqrt{\mu_2}}
\right)^3
$$
where $\mu_1 = EX$ and $\mu_2 = \mbox{Var}(X)$. The corresponding sample estimate
skewness $\Gamma_{3,N}$ is given by
$$
\Gamma_{3,N}
\stackrel{\Delta}{=}
\dfrac{1}{N} \sum_{i=1}^{N} \left(
\dfrac{x_i - m_N}{s_N}
\right)^3
$$ 
where $m_N$ and $s_N$ are the sample mean and standard deviation respectively.
By the strong law of large number $\Gamma_{3,N} \rightarrow \gamma_3$ as $N \rightarrow \infty$.
Clearly if $f_X()$ is symmetric then $\gamma_3 \equiv 0$, so that
$\left|\Gamma_{3,N}\right|$
can be used as a test statistic since its large values betoken asymmetry.
The critical values for this test come from the asymptotic noramlity 
of $\Gamma_{3,N}$.
It is shown on pages 386--387 of @lehmannlargesample that
$$
\sqrt{N}\,(\Gamma_{3,N} - \gamma_3) \rightarrow N(0,b)\;\;
\mbox{where}\;\;
b \stackrel{\Delta}{=}
\dfrac{
\mu_6 - 6 \mu_2 \mu_4 - \mu_3^2 + 9 \mu_2^3 
}
{\mu_2^3}
$$
where $\mu_k \stackrel{\Delta}{=} E\left(X - EX)\right)^k\;$ for $k \in \{2,3,4,5,6\}$.

***


# A different test of symmetry <a name="newtest"></a> #
Let $f_X$ and $F_X$ be the population probability density function and cumulative
distribution function respectively.
Considering the (potential) heaviness of the tails of $f_X$, we can only reasonably
assume boundedness of the first moment
$\int_{\mathbb{R}} \left|r \right| f_X(r)\,dr < \infty$. That is, the expectation
$EX$ is finite. We propose using as the test statistic the absolute value of the difference between
the sample mean and sample median normalized by an indication of the _scale_ parameter;
that is, $T_N$ as defined below:

$$
T_N \stackrel{\Delta}{=}
\dfrac{\left|m_N  - \zeta_{1/2,N}\right|}
{R_N}
\;\;\mbox{where}\;\;
R_N \stackrel{\Delta}{=} \dfrac{\zeta_{3/4,N} - \zeta_{1/4,N}}
{2 \times 0.6745}
$$
where by $\zeta_{p,N}$ we mean the the $p$'th sample quantile.
Our choice of $R_N$ is motivated by the discussion on page 86 of @serfling where
in particular the strange looking denominator of $R_N$ is explained.
Large values of $T_N$ suggest rejecting the null hypothesis:
$$ H_0 : \;\;f_X\;\;\mbox{is symmetric}$$

Clearly we need to approximate the distribution of $T_N$ under $H_0$; to this
end we use a trick suggested by @hinkley and proceed as follows:
Let $F_N$
be the empirical cumulative distribution function of the data  $x = [x_1, x_2,\ldots,
x_N]$. By the authority of the Gilvenko-Cantelli theorem [@probandmeasure page 269],
$F_N$ is an approximation of $F_X$.
Create a _new_ data set $y = [y_1, y_2,\ldots,y_N]$ by setting
$$ y_i \stackrel{\Delta}{=} 2\,\zeta_{1/2,N} - x_i \;\;\mbox{for}\;\; i \in \{1,\ldots,N\}$$
and let $G_N$ be the empirical cdf of $y$.
If the null hypothesis $H_0$ is true, then $G_N$ is also an approximation of $F_X$.
Consequently, under $H_0$, the cumulative distribution function $K_N$ defined by
$$ K_N = \dfrac{1}{2} F_N + \dfrac{1}{2} G_N $$
is also an approximation of $F_X$. Observe that $K_N$ is the empirical CDF
of the data vector $w$ that we obtain by concatenating the vectors $x$ and $y$
as $w \stackrel{\Delta}{=} [x_1,\ldots,x_N, y_1,\ldots, y_N]$.
Clearly $K_N$ is a symmetric distribution.
We can therefore approximate the null distribution of our test statistic $T_N$
by forming bootstrap samples
$$w_1^\ast, w_2^\ast,\ldots,w_B^\ast$$
from $w$ 
and applying $T_N$ to each bootstrap sample to obtain
$$T_{N,1}^\ast, T_{N,2}^\ast,\ldots,
T_{N,B}^\ast$$
The corresponding _basic bootstrap_ $p$-value for $T_N$ is given by

$$ p_N^\ast \stackrel{\Delta}{=} \dfrac{1 + \sum_{i=1}^{B} \mathbb{1}\left(
T_{N,i}^\ast \geq T_N
\right)}
{1 + B}
$$ 
The above basic bootstrap estimate can be improved upon by using antithetic variance reduction
and balanced bootstrap, as the following implementation in [`R 3.2.0`](http://www.R-project.org)
[@theRlang]
does.

```{r}
rm (list = ls())
require (gtools)

lu <- defmacro (j, k, expr={
  c((k*(j-1)+1):(k*j))
})

symmetry.statfunc <- function (y) {
  mu <- mean (y);
  md <- median (y);
  tmp <- quantile (x=y, probs=c(1/4, 3/4));
  s <- (tmp[2] - tmp[1])/ (2 * 0.6745);  # Indication of scale
  val <- abs (mu - md) / s;
  names (val) <- NULL
  val;
}

symmetryTest <- function (x, B = 1000) {
  stopifnot (class(x) == "numeric")
  result <- list();
  result$statistic <- symmetry.statfunc (x);
  result$mu <- mean(x);
  result$md <- median(x);
  tmp <- quantile (x=x, probs=c(1/4, 3/4));
  result$scale  <- (tmp[2] - tmp[1])/ (2 * 0.6745);  # Indication of scale
  names (result$scale) <- NULL
  
  Hnull = c(x, (2*median(x)-x)) # This is the symmetrized data which is our proxy for H0
  
  # Note the use of balanced as opposed to ordinary bootstrap,
  # together with the use of antithetic variates.
  # Page 493 of Hinkley et al. exercise 21
  halfB = as.integer (B/2);
  direct.val = rep (0.0, halfB)
  anti.val = rep (0.0, halfB)
  
  BI <- rep (c(1:length(Hnull)), halfB);
  II <- sample (x=BI, size=length(BI), replace=FALSE)
  
  for (i in 1:halfB) {
    idx = lu (i, length(Hnull))
    jdx = II[idx]
    direct.val[i] = symmetry.statfunc(Hnull[jdx])
    
    jdx = (length(Hnull)+1) - jdx
    anti.val[i] = symmetry.statfunc (Hnull[jdx])
  }
  
  val = c (direct.val, anti.val)
  result$p.val <- (1 + sum (val >= result$statistic))/(1+B);
 
  result
}
```

***

## Experimenting with known distribution <a name="experimenting"></a> ##
To gauge the performance of [the above procedure](#newtest) we first try it on simulated data
with
known asymmetry (or symmetry). The following plot shows the pdf of the stable
law with characteristic index $\alpha = 1.4$ 
and skew parameter $\beta=0.45$ as well as the symmetric stable law with $\alpha=1.4$ (i.e. $\beta=0$).
Neither distribution has
finite variance (A reference on stable laws is @bigstablebook).

<a name="stablesfig"></a>
```{r, fig.width = 7, fig.height=5, echo=FALSE}
library (stabledist)
alpha = 1.4
beta = 0.45
mymar <- par()$mar
mymar[2] <- 1.2 * mymar[2]
par (mar=mymar)
x = seq (from=-5, to=5, length=500)
u = dstable (x, alpha=alpha, beta=0.0)
v = dstable (x, alpha=alpha, beta=beta)
plot (x, u, type='l', col='slateblue4',
ylab="probability density",
main="Stable distribution alpha=1.4 with and w/o skewness"); grid()
lines (x, v, type='l', col='red')
legend ("topleft", legend=c('beta=0', sprintf("beta=%.2f", beta)),
lty = c(1,1), col=c('slateblue4', 'red'), pch = c(NA,NA))
```

We simulate samples corresponding to the above two distribution and subject them to our [test](#newtest).
```{r}
set.seed (35384)
data = rstable (n=300, alpha=alpha, beta=0.0)
res = symmetryTest (data, B=2000)
print (res$p.val)
```
As expected, there is no evidence against symmetry for the blue density.

```{r}
set.seed (35384)
print (beta)
data = rstable (n=300, alpha=alpha, beta=beta)
res = symmetryTest (data, B=2000)
print (res$p.val)
```
As expected, the test rejects $H_0$ for the red density
in the [above figure](#stablesfig).

Next we apply the [test](#newtest) to data simulated from $t$ distribution
with $\mbox{df}=2.1$, with non-centrality parameter $\mbox{nc}=0$ and
$\mbox{nc}=0.45$, which are plotted below. Both $t_{\mbox{df}=2.1}$ and
$t_{\mbox{df}=2.1, \mbox{nc}=0.45}$ have unbounded third moments. Note
that we have centred the non-central $t$ pdf at its mean so as to highlight
the difference of skewnesses of the two pdf's.

<a name="students"></a>
```{r, fig.width=7, fig.height=5, echo=FALSE}
meannct <- function (degfreedom, noncentrality) {
  a = noncentrality * sqrt (degfreedom/2)
  b = lgamma ((degfreedom - 1)/2) - lgamma (degfreedom/2)
  val = a * exp(b)
  val
}

df = 2.1
ncp = 0.45

u = dt (x, df=df, ncp=0.0)
v = dt (x, df=df, ncp=ncp)
plot (x, u, type='l', col='slateblue4',
ylab="probability density",
main="t distribution df=2.1 with and w/o non-centrality"); grid()
lines ((x-meannct(df, ncp)), v, type='l', col='red')
legend ("topleft", legend=c('ncp=0, mean=0', sprintf("ncp=%.2f, mean=0", ncp)),
lty = c(1,1), col=c('slateblue4', 'red'), pch = c(NA,NA))
```

```{r}
set.seed (35384)
print (df)
data = rt (n=300, df=df)
print (symmetryTest (data)$p.val)
```
As expected there is no evidence of asymmetry for the blue curve.
```{r}
set.seed (35384)
print (ncp)
data = rt (n=300, df=df, ncp=ncp)
print (symmetryTest (data)$p.val)
```
But the test rejects the red curve.

***

# Trying the test out on market data <a name="marketdata"></a>#
We use the [quantmod](http://CRAN.R-project.org/package=quantmod) [@quantmod] package of R 3.2.0 to
download market data from [Yahoo Finance Canada](https://ca.finance.yahoo.com/).
```{r}
rm (list = ls())
library (quantmod)
source ("./getTicker.R") # This is a just a convenient wrapper for getSymbols{quantmod}
source ("symmetryTest.R") 
set.seed (38535)
```
Let us try the [S&P 500](https://en.wikipedia.org/wiki/S&P_500) daily log-returns
for the past two years.
```{r}
SNP500 <- getTicker (sym="^GSPC", freq="daily", from="2013-07-10", to="2015-07-10")
SNP500$symTest = symmetryTest (SNP500$rt, B=2000)
print (SNP500$symTest$p.val)
```
There is no evidence of asymmetry in S&P 500 returns.
Next we try [PowerShares QQQ Trust, Series 1](https://ca.finance.yahoo.com/q?s=QQQ&ql=1)
daily log-returns for the same period.
```{r}
getTicker (sym="QQQ", auto.assign=TRUE,
           freq="daily", from="2013-07-10", to="2015-07-10")
QQQ$symTest = symmetryTest (QQQ$rt, B=2000)
print (QQQ$symTest$p.val)
sprintf ("QQQ daily log returns: mean=%.4f, median=%.4f", QQQ$symTest$mu, QQQ$symTest$md) 
```
There is evidence of asymmetry in
[QQQ](https://ca.finance.yahoo.com/q?s=QQQ&ql=1) daily log-returns, and the claim of negative skewness
made in
the cited [article](http://www.bloombergview.com/articles/2015-06-25/fear-factor-drives-the-stock-market-s-highs-and-lows) is verified since the mean is less than the median.
However, there does not appear to be any strong indication of negative skewness of daily log-returns 
of IBM and Microsoft for the same period:
```{r}
getTicker (sym="IBM", auto.assign=TRUE,
           freq="daily", from="2013-07-10", to="2015-07-10")
IBM$symTest = symmetryTest (IBM$rt, B=2000)
print (IBM$symTest$p.val)

getTicker (sym="MSFT", auto.assign=TRUE,
           freq="daily", from="2013-07-10", to="2015-07-10")
MSFT$symTest = symmetryTest (MSFT$rt, B=2000)
print (MSFT$symTest$p.val)
```
For the [S&P/TSX Composite Index](https://en.wikipedia.org/wiki/S&P/TSX_Composite_Index)
of the [Toronto Stock Exchange](https://en.wikipedia.org/wiki/Toronto_Stock_Exchange),
there appears to be a very strong evidence of negative skewness since the p-value
is small and the mean is less than the median:
```{r}
TSE = getTicker (sym="^GSPTSE",
           freq="daily", from="2013-07-10", to="2015-07-10")
TSE$symTest = symmetryTest (TSE$rt, B=2000)
print (TSE$symTest$p.val)
sprintf ("TSX daily log returns: mean=%.4f, median=%.4f", TSE$symTest$mu, TSE$symTest$md) 
```

***

# Conclusion #
Presence of asymmetry and negative skewness in the distribution of the log-returns
of stocks and stocks indices is not necessarily universally true, and should be
tested on a case-by-case basis.

***

# References #



